Calendar-based graphics for visualising people's daily schedules

### 2017-07-14 main.Rmd
added more to Section 4

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  It is challenging to visualise people's daily schedules as such data are often of multiple seasons and of long time span. One example is Melbourne pedestrian sensor data which are collected at hourly interval and at a number of locations. Owing to human activities, the data features multiple temporal scales such as time of day, day of week, day of year. We develop the \code{frame\_calendar} function that provides a method for rearranging such kind of data in a calendar-based format and accomplishing visualisation with the grammar of graphics. We describe its construction along with a few options. We provide some examples to demonstrate its usage by applying to the pedetrian data.
keywords:
  # at least one keyword must be supplied
  formatted: [calendar-based visualisation, temporal-context data]
  plain:     [calendar-based visualisation, temporal-context data]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

Calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context. For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It remarkably contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colours encoded as quantities, it may become less perceivable of the actual variations or behaviours from the viewer's perspective.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is the implementation of the grammar of graphics (@wilkinson2006grammar and @wickham2010layered).

@@

When \code{polar = TRUE}, the polar transformation is carried out for the data.

```{r flinders-polar, fig.height = 6.5, fig.cap = "Line gplyphs on polar coordinates."}
```

@@

As

```{r facet, fig.height = 11, fig.cap = "Line graphs faceted by the 3 sensors in the monthly calendar."}
```

@@

The \code{frame_calendar} function does not constrain itself to mapping only temporal variable to the \code{x}. Figure \ref{fig:scatterplot} shows the lag scatterplot where hourly counts are assigned to the \code{x} and lagged themselves to the \code{y}. It provides a visual tool for identifying repeating patterns.

```{r scatterplot, fig.height = 5, fig.cap = "Lag scatterplot in the daily calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at lag 1. The correlation between them is more consitent on non-work days than work days."}
```

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  It is challenging to visualise people's daily schedules as such data are often of multiple seasons and of long time span. One example is Melbourne pedestrian sensor data which are collected at hourly interval and at a number of locations. Owing to human activities, the data features multiple temporal scales such as time of day, day of week, day of year. We develop the \code{frame\_calendar} function that provides a method for rearranging such kind of data in a calendar-based format and accomplishing visualisation with the grammar of graphics. We describe its construction along with a few options. We provide some variations to demonstrate its usage by applying to the pedetrian data over one-year period of 2016.
keywords:
  # at least one keyword must be supplied
  formatted: [calendar-based visualisation, temporal-context data]
  plain:     [calendar-based visualisation, temporal-context data]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

Calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context. For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It remarkably contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using coloured blocks rather than curves, it may become perceptually difficult to accurately judge the shape positions and changes, although the glyphs come with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is the implementation of the grammar of graphics [@wilkinson2006grammar; @wickham2010layered].

@@

When \code{polar = TRUE}, the polar transformation is carried out for the data. Again the computation remain similar to the one described in @Wickham2012glyph. Figure \ref{fig:flinders-polar} is considered as the spiral display of Figure \ref{fig:flinders-2016}.

```{r flinders-polar, fig.height = 6.5, fig.cap = "Line glyphs on polar coordinates."}
```

@@

Placing multiple series on the common scale makes the magnitudes comparable between them; the glyphs are yet small in size, resulting in the overlapping problem. To avoid the problem, the calendar layout can be embedded into a series of plots for the different sensors. Figure \ref{fig:facet} presents the idea of the graphs of calendar plots. By doing so, the comparability with respect to the magnitudes is sacrificed for the emphasis of the individual sensor.

```{r facet, fig.height = 11, fig.cap = "Line graphs faceted by the 3 sensors in the 6 by 2 monthly calendar."}
```

@@

The \code{frame_calendar} function does not constrain itself to mapping only temporal variable to the \code{x}. Figure \ref{fig:scatterplot} shows the lag scatterplot, where the current hourly count is assigned to the \code{x} and the lagged hourly count to the \code{y} at Flinders Street Station, organised in the daily calendar. It provides a visual tool for identifying repeating patterns. Figure \ref{fig:scatterplot} indicates two separate paths in the work-day glyphs, which suggests that an hour of a work day with many pedestrians is likely to be followed by the time of either more pedestrians or fewer. Quite the contrary, the relationship is so consistent on the non-work days. This is a clear sign of bimodalily on work days whereas unimodality on the rest of the days, which is also supported by Figure \ref{fig:flinders-2016}.

```{r scatterplot, fig.height = 5, fig.cap = "Lag scatterplot in the daily calendar layout. Previous hour's count is plotted against each hour's count at Flinders Street Station to demonstrate the autocorrelation at lag 1. The correlation between them is more consitent on non-work days than work days."}
```

### 2017-07-14 main.tex
added more to Section 4

\Abstract{
It is challenging to visualise people's daily schedules as such data are
often of multiple seasons and of long time span. One example is
Melbourne pedestrian sensor data which are collected at hourly interval
and at a number of locations. Owing to human activities, the data
features multiple temporal scales such as time of day, day of week, day
of year. We develop the \code{frame\_calendar} function that provides a
method for rearranging such kind of data in a calendar-based format and
accomplishing visualisation with the grammar of graphics. We describe
its construction along with a few options. We provide some examples to
demonstrate its usage by applying to the pedetrian data.
}

@@

Calendar-based graphics turn out to be a useful tool in unfolding
human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It remarkably
contrasts weekdays and weekends, highlights public holiday, and presents
other known seasons like school vacations, all of which have influence
over the turn-outs in the office. The calendar-based heatmap was
implemented in a couple of R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are too constrained to colour-encoding graphics and day of
week as the smallest time scale. Time of day, which serves as one of the
most important aspect in explaining variations arisen from pedestrian
sensor data, will be neglected through daily aggregation. Additionally
if simply using colours encoded as quantities, it may become less
perceivable of the actual variations or behaviours from the viewer's
perspective.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is the
implementation of the grammar of graphics (\citet{wilkinson2006grammar}
and \citet{wickham2010layered}).

@@

When \code{polar = TRUE}, the polar transformation is carried out for
the data.

@@

\caption[Line gplyphs on polar coordinates]{Line gplyphs on polar coordinates.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}

@@

As

@@

\caption[Line graphs faceted by the 3 sensors in the monthly calendar]{Line graphs faceted by the 3 sensors in the monthly calendar.}\label{fig:facet}
\end{figure}
\end{CodeChunk}

@@

The \code{frame_calendar} function does not constrain itself to mapping
only temporal variable to the \code{x}. Figure \ref{fig:scatterplot}
shows the lag scatterplot where hourly counts are assigned to the
\code{x} and lagged themselves to the \code{y}. It provides a visual
tool for identifying repeating patterns.

@@

\caption[Lag scatterplot in the daily calendar]{Lag scatterplot in the daily calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at lag 1. The correlation between them is more consitent on non-work days than work days.}\label{fig:scatterplot}
\end{figure}
\end{CodeChunk}

>>>>

\Abstract{
It is challenging to visualise people's daily schedules as such data are
often of multiple seasons and of long time span. One example is
Melbourne pedestrian sensor data which are collected at hourly interval
and at a number of locations. Owing to human activities, the data
features multiple temporal scales such as time of day, day of week, day
of year. We develop the \code{frame\_calendar} function that provides a
method for rearranging such kind of data in a calendar-based format and
accomplishing visualisation with the grammar of graphics. We describe
its construction along with a few options. We provide some variations to
demonstrate its usage by applying to the pedetrian data over one-year
period of 2016.
}

@@

Calendar-based graphics turn out to be a useful tool in unfolding
human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It remarkably
contrasts weekdays and weekends, highlights public holiday, and presents
other known seasons like school vacations, all of which have influence
over the turn-outs in the office. The calendar-based heatmap was
implemented in a couple of R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are too constrained to colour-encoding graphics and day of
week as the smallest time scale. Time of day, which serves as one of the
most important aspect in explaining variations arisen from pedestrian
sensor data, will be neglected through daily aggregation. Additionally
if simply using coloured blocks rather than curves, it may become
perceptually difficult to accurately judge the shape positions and
changes, although the glyphs come with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is the
implementation of the grammar of graphics
\citep{wilkinson2006grammar, wickham2010layered}.

@@

When \code{polar = TRUE}, the polar transformation is carried out for
the data. Again the computation remain similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:flinders-polar} is considered
as the spiral display of Figure \ref{fig:flinders-2016}.

@@

\caption[Line glyphs on polar coordinates]{Line glyphs on polar coordinates.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}

@@

Placing multiple series on the common scale makes the magnitudes
comparable between them; the glyphs are yet small in size, resulting in
the overlapping problem. To avoid the problem, the calendar layout can
be embedded into a series of plots for the different sensors. Figure
\ref{fig:facet} presents the idea of the graphs of calendar plots. By
doing so, the comparability with respect to the magnitudes is sacrificed
for the emphasis of the individual sensor.

@@

\caption[Line graphs faceted by the 3 sensors in the 6 by 2 monthly calendar]{Line graphs faceted by the 3 sensors in the 6 by 2 monthly calendar.}\label{fig:facet}
\end{figure}
\end{CodeChunk}

@@

The \code{frame_calendar} function does not constrain itself to mapping
only temporal variable to the \code{x}. Figure \ref{fig:scatterplot}
shows the lag scatterplot, where the current hourly count is assigned to
the \code{x} and the lagged hourly count to the \code{y} at Flinders
Street Station, organised in the daily calendar. It provides a visual
tool for identifying repeating patterns. Figure \ref{fig:scatterplot}
indicates two separate paths in the work-day glyphs, which suggests that
an hour of a work day with many pedestrians is likely to be followed by
the time of either more pedestrians or fewer. Quite the contrary, the
relationship is so consistent on the non-work days. This is a clear sign
of bimodalily on work days whereas unimodality on the rest of the days,
which is also supported by Figure \ref{fig:flinders-2016}.

@@

\caption[Lag scatterplot in the daily calendar layout]{Lag scatterplot in the daily calendar layout. Previous hour's count is plotted against each hour's count at Flinders Street Station to demonstrate the autocorrelation at lag 1. The correlation between them is more consitent on non-work days than work days.}\label{fig:scatterplot}
\end{figure}
\end{CodeChunk}
